Abstractions for Type-Level Programming

### 2022-03-18 article.tex
Slides

This chapter presents an alternative construct for type-level programming based on subtyping, which we call \emph{match types}.
As the name suggests, match types allow programmers to express types that perform pattern matching on types:
%
\begin{lstlisting}
type Elem[X] = X match {
  case String => Char
  case List[t] => Elem[t]
  case Any => X
}
\end{lstlisting}
%
The example, which we explain in detail in \Cref{sec:overview}, defines the type |Elem| by matching on the type parameter |X|.
%
We have implemented match types in the latest version of Scala, an industry-grade, production-ready compiler.
Match types have received a great interest from the Scala community, and are already in active use.

@@

In this section, we offer a brief introduction of match types in Scala by inspecting the example from the previous section in more detail:
%
\begin{lstlisting}
type Elem[X] = X match {
  case String => Char
  case List[t] => Elem[t]
  case Any => X
}
\end{lstlisting}
%
This example defines a type |Elem| parametrized by one type parameter |X|. The right-hand side is defined in terms of a match on the type parameter -- a \emph{match type}.
A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
For example, the above type reduces as follows:
%
\begin{lstlisting}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[List[Int]] =:= Int
\end{lstlisting}
%
Here we use |S =:= T| to denote type equality between the two types |S| and |T|, witnessed by mutual subtyping.
To reduce a match type, the scrutinee is compared to each pattern, one after the other, using \emph{subtyping}.
For example, although |String| is a subtype of both |String| and |Any| (the top of Scala's subtyping lattice), |Elem[String]| reduces to |Char| because the corresponding case appears first.

@@

Match types enable a lightweight form of dependent typing, since term-level pattern matching expressions can be typed accordingly at the type level as a match types.
Consider the following function definition:
%
\begin{lstlisting}
def elem[X](x: X): Elem[X] = x match {
  case x: String => x.charAt(0)
  case x: List[t] => elem(x.head)
  case x: Any => x
}
\end{lstlisting}
%
This definition is well-typed because the match expression in |elem|'s body has the exact same scrutinee and pattern types as |Elem[X]| (the function's return type).

@@

To show why empty types are problematic, we can construct an example where, in the absence of an inhabitance check, the same match type could be reduced differently in two different contexts:
%
\begin{lstlisting}
type M[X] = X match {
  case Int => String
  case String => Int
}
class C {
  type X
  def f(bad: M[X & String]): Int = bad
}
class D extends C {
  type X = Int
}
\end{lstlisting}
%
In this example, the definition of |f| in |C| type-checks because |X| |&| |String| and |Int| are disjoint (since |String| and |Int| are disjoint) and |M[X| |&| |String]| reduces to |Int| (|M|'s second case applies).
Class |D| refines the definition of |C| by giving concrete definition of |X|.
The unsoundness manifests itself in the body of class |D|, where |X| |&| |String| is a subtype of |Int| and |M[X| |&| |String]| reduces to |String| (|M|'s first case applies).
There, it is possible to call the function~|f| with a string argument, which would result in a runtime error.
Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent |M[Int| |&| |String]| from reducing given that |Int| |&| |String| is not inhabited.

@@

In their introductory example,~\citeauthor{weirich2011generative} show how type family reduction can lead to unsoundness in the absence of role annotations.
Their presentation also includes a hypothetical translation of their example to Standard ML, which translates directly to Scala as follows:
%
\begin{lstlisting}
trait AgeClass {
  type Age
  def addAge(a: Age, i: Int): Int
}
object AgeObject extends AgeClass {
  type Age = Int
  def addAge(a: Age, i: Int): Int = a + i
}
\end{lstlisting}
%
In this example, type |Age| is abstract in |AgeClass| and concrete in |AgeObject|.
In the pre-role Haskell equivalent of this example, the unsoundness comes when the above definitions are combined with a type family that discriminates |Age| and |Int|.
Such type family would reduce differently in |AgeClass|, where the types are different, and in |AgeObject|, where those two types are synonyms, which can easily be exploited to obtain a runtime error.

Luckily, our design of match types is not affected by this issue.
The reason comes from the use of subtyping (and disjointness), which shields our implementation from incorrectly discriminating |Age| from |Int| when |Age| is abstract.
Consider the following match type definition (directly translated from the Haskell example):
%
\begin{lstlisting}
type M[X] = X match {
  case Age => Char
  case Int => Bool
}
\end{lstlisting}
%
Our algorithm would not reduce |M[Int]| to |Bool| in |AgeClass| as this reduction would require evidence that |Age| and |Int| are disjoint, which cannot be constructed when |Age| is an unbounded abstract type.

>>>>

This chapter presents an alternative construct for type-level programming based on subtyping, which we call \emph{match types}.
As the name suggests, match types allow programmers to express types that perform pattern matching on types:
%
\begin{lstlisting}
type Elem[X] = X match
  case String => Char
  case List[t] => Elem[t]
  case Any => X
\end{lstlisting}
%
The example, which we explain in detail in \Cref{sec:overview}, defines the type |Elem| by matching on the type parameter |X|.
%
We have implemented match types in the latest version of Scala, an industry-grade, production-ready compiler.
Match types have received a great interest from the Scala community, and are already in active use.

@@

In this section, we offer a brief introduction of match types in Scala by inspecting the example from the previous section in more detail:
%
\begin{lstlisting}
type Elem[X] = X match
  case String => Char
  case List[t] => Elem[t]
  case Any => X
\end{lstlisting}
%
This example defines a type |Elem| parametrized by one type parameter |X|. The right-hand side is defined in terms of a match on the type parameter -- a \emph{match type}.
A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
For example, the above type reduces as follows:
%
\begin{lstlisting}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[List[Int]] =:= Int
\end{lstlisting}
%
Here we use |S =:= T| to denote type equality between the two types |S| and |T|, witnessed by mutual subtyping.
To reduce a match type, the scrutinee is compared to each pattern, one after the other, using \emph{subtyping}.
For example, although |String| is a subtype of both |String| and |Any| (the top of Scala's subtyping lattice), |Elem[String]| reduces to |Char| because the corresponding case appears first.

@@

Match types enable a lightweight form of dependent typing, since term-level pattern matching expressions can be typed accordingly at the type level as a match types.
Consider the following function definition:
%
\begin{lstlisting}
def elem[X <: Singleton](x: X): Elem[X] = x match
  case x: String => x.charAt(0)
  case x: List[t] => elem(x.head)
  case x: Any => x
\end{lstlisting}
%
This definition is well-typed because the match expression in |elem|'s body has the exact same scrutinee and pattern types as |Elem[X]| (the function's return type).

@@

To show why empty types are problematic, we can construct an example where, in the absence of an inhabitance check, the same match type could be reduced differently in two different contexts:
%
\begin{lstlisting}
type M[X] = X match
  case Int => String
  case String => Int
class C:
  type X
  def f(bad: M[X & String]): Int = bad
class D extends C:
  type X = Int
\end{lstlisting}
%
In this example, the definition of |f| in |C| type-checks because |X| |&| |String| and |Int| are disjoint (since |String| and |Int| are disjoint) and |M[X| |&| |String]| reduces to |Int| (|M|'s second case applies).
Class |D| refines the definition of |C| by giving concrete definition of |X|.
The unsoundness manifests itself in the body of class |D|, where |X| |&| |String| is a subtype of |Int| and |M[X| |&| |String]| reduces to |String| (|M|'s first case applies).
There, it is possible to call the function~|f| with a string argument, which would result in a runtime error.
Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent |M[Int| |&| |String]| from reducing given that |Int| |&| |String| is not inhabited.

@@

In their introductory example,~\citeauthor{weirich2011generative} show how type family reduction can lead to unsoundness in the absence of role annotations.
Their presentation also includes a hypothetical translation of their example to Standard ML, which translates directly to Scala as follows:
%
\begin{lstlisting}
trait AgeClass:
  type Age
  def addAge(a: Age, i: Int): Int
object AgeObject extends AgeClass:
  type Age = Int
  def addAge(a: Age, i: Int): Int = a + i
\end{lstlisting}
%
In this example, type |Age| is abstract in |AgeClass| and concrete in |AgeObject|.
In the pre-role Haskell equivalent of this example, the unsoundness comes when the above definitions are combined with a type family that discriminates |Age| and |Int|.
Such type family would reduce differently in |AgeClass|, where the types are different, and in |AgeObject|, where those two types are synonyms, which can easily be exploited to obtain a runtime error.

Luckily, our design of match types is not affected by this issue.
The reason comes from the use of subtyping (and disjointness), which shields our implementation from incorrectly discriminating |Age| from |Int| when |Age| is abstract.
Consider the following match type definition (directly translated from the Haskell example):
%
\begin{lstlisting}
type M[X] = X match
  case Age => Char
  case Int => Bool
\end{lstlisting}
%
Our algorithm would not reduce |M[Int]| to |Bool| in |AgeClass| as this reduction would require evidence that |Age| and |Int| are disjoint, which cannot be constructed when |Age| is an unbounded abstract type.

### 2022-03-18 preprint.tex
Add footnote and QE and character class
Draft submission
Slides

Our extension does not model virtual dispatch explicitly in singleton types.
Instead, the result type of a method call |t.m(|\ldots{}|)| is always the result type of |m| in |t|'s static type.
Consequently, |dependent| methods effectively become |final|, given that only a provably-equivalent implementation could be used to override it.

@@

\section{Use Case}
\label{sec:use-case}

@@

\subsection{A Type-Safe Database Interface}

@@

>>>>

Our extension does not model virtual dispatch explicitly in singleton types.
Instead, the result type of a method call |t.m(|\ldots{}|)| is always the result type of |m| in |t|'s static type.
Consequently, |dependent| methods effectively become |final|, given that only a provably-equivalent implementation could be used to override it, modulo side effects.

@@

\section{Case Study: A Type-Safe Database Interface}
\label{sec:use-case}

@@

\subsection{Type-Safe Datasets}

@@

% TODO: Compare to pure subtype systems

### 2022-03-18 slides.tex
Slides

\usepackage[outputdir=latex.out,]{minted}
\usepackage{fontspec}

%% Minted setup (syntax highlighting)
\usemintedstyle{vs}
\setminted{fontsize=\footnotesize}
\newminted[code]{scala}{gobble=0}
\newenvironment{slide}[2][]
  {\begin{frame}[fragile,environment=slide,#1]{#2}}
  {\end{frame}}
\newcommand\makemintedshortinline[2]{%
  \catcode`#2=13
  \begingroup%
  \lccode`\~=`#2
  \lowercase{\endgroup\protected\def~{\mintinline{#1}~}}%
}

@@

@@

@@

\begin{frame}{Table of contents}
  \setbeamertemplate{section in toc}[sections numbered]
  \tableofcontents
\end{frame}

\section[Type-Level Programming\\What? Why?]{What? Why?}

\begin{slide}{Example 1: Database queries (2016)}
Spark API, practically untyped:
\begin{code}
phoneBook                                           : DataFrame
  .innerJoin(cityMap, col("address"))               : DataFrame
  .select(col("name"), col("neighborhood"))         : DataFrame
  .filter(x => uniqueName(x.getAs[String]("name"))) : DataFrame
  .groupBy(col("neighborhood")).count()             : DataFrame
  .sortDesc(col("count"))                           : DataFrame
  .head.getAs[String](0)                            : String
\end{code}
\pause
The same API, revisited with type-level programming:
\begin{code}
phoneBook                          : DF[PhoneBookEntry]
  .innerJoin(cityMap, 'address)    : DF[(Address, String, String, ... )]
  .select('_2, '_3).as[Family]()   : DF[Family]
  .filter(x => uniqueName(x.name)) : DF[Family]
  .groupBy('neighborhood).count()  : DF[(String, Long)]
  .sortDesc('_2)                   : DF[(String, Long)]
  .head._1                         : String
\end{code}
\end{slide}

\begin{slide}{Example 2: Regular expressions (2022)}
Regular expressions in Scala's standard library:
\begin{code}
val rational = new Regex("""(\d+)(?:\.(\d+))?""")
"3.1415" match
  case rational(captures: Seq[String]) =>
    val i = captures(0)
    val j = captures(1)
\end{code}
\pause
The same API, revisited with type-level programming:
\begin{code}
val rational = new R("""(\d+)(?:\.(\d+))?""")
"3.1415" match
  case rational(captures: (String, Option[String])) =>
    val i = captures._1
    val j = captures._2
\end{code}
\end{slide}

\begin{slide}{What changed between 2016 and 2022?}
\begin{itemize}
  \item Example 1 uses a hack, implicits:
  \begin{itemize}
    \item convoluted to use
    \item slow to compile
    \item popular nonetheless
  \end{itemize}
  \pause
  \item Example 2 uses a first-class language construct, match types:
  \begin{itemize}
    \item easier to use
    \item faster to compile
    \item solid theoretical foundation
  \end{itemize}
\end{itemize}
\end{slide}

@@

\begin{slide}{Compilation times for the type-level join operation, zoomed in}
\begin{center}
\input{figures/join-2.gex}
\end{center}
\end{slide}

\section{Generalizing Scala's Singleton~Types}

\begin{slide}{Singleton types in Scala}
Scala supports a few singleton types:
\begin{itemize}
  \item |x.type|, the type of the variable |x| (since forever)
  \item |42|, the type of the integer literal |42| (since 2016)
  \item |+|, the type of the primitive addition (since 2020)
\end{itemize}
\pause
How much of the language can we lift to the type level?
\end{slide}

\begin{slide}{Generalized singleton types: a proposal}
\end{slide}

\section{Type-Level Programming with Match Types}

\end{document}

>>>>

\usepackage{fontspec}

%% Minted setup (syntax highlighting)
\usepackage[outputdir=latex.out]{minted}
\usemintedstyle{vs}
\setminted{fontsize=\footnotesize}
\newminted[code]{scala}{gobble=0, escapeinside=@@}
\newenvironment{slide}[2][]
  {\begin{frame}[fragile,environment=slide,#1]{#2}}
  {\end{frame}}
\newcommand\makemintedshortinline[2]{%
  \catcode`#2=13
  \begingroup%
  \lccode`\~=`#2
  \lowercase{\endgroup\protected\def~{\mintinline{#1}~}}%
}
\renewcommand{\fcolorbox}[4][]{#4}

@@

\usepackage{xspace}
\usepackage{mathtools}
\def\C{\normalfont\text{C}}
\def\<:{\allowbreak\operatorname{\normalfont\text{<}:}\allowbreak}
\def\Fsub{\normalfont{F}\ensuremath{_{\<:}}\xspace}

\usepackage[autostyle=false, style=english]{csquotes}
\MakeOuterQuote{"}

\usepackage{xcolor}
\definecolor{typeColor}{HTML}{2B91B0}
\definecolor{implColor}{HTML}{57B5E8}
\definecolor{singColor}{HTML}{009E73}
\definecolor{mtpeColor}{HTML}{9400D4}
\newcommand{\tp}{\color{typeColor}}

\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,positioning, arrows.meta}

@@

\newcommand{\timeline}[5] {
  \begin{tikzpicture}
    \draw[lightgray!#5!black, thick, -Triangle] (0,0) -- (10cm,0);
    \foreach \x in {0,1,...,6}
    \draw[lightgray!#5!black] (\x*1.4 cm,3pt) -- (\x*1.4 cm,-3pt);

    \foreach \x/\descr in {0/2016,1/2017,2/2018,3/2019,4/2020,5/2021,6/2022}
    \node[lightgray!#5!black, font=\scriptsize, text height=1.75ex,
    text depth=.5ex] at (\x*1.4,-.3) {$\descr$};

    \fill[lightgray!#1!implColor] (0.25, 0.5) circle (4pt) node[lightgray!#1!black,align=center,above=7pt,font=\small]{DataFrame\\Library};

    \draw[lightgray!#2!singColor, line width=4pt]
    (2.25*1.4,.5) -- (4.417*1.4,.5) node[lightgray!#2!black,midway,above=5pt,font=\small]{Generalized Singletons};

    \draw[lightgray!#3!mtpeColor, line width=4pt]
    (2.5*1.4,-.7) -- (6.15*1.4,-.7) node[lightgray!#3!black,midway,below=5pt,font=\small]{Match Types};
    \draw[thick,lightgray!#3!mtpeColor,-Triangle,line width=1.2pt] (6.15*1.4,-.7) -- (6.25*1.4,-.7);

    \fill[lightgray!#4!mtpeColor] (6.1*1.4, 0.5) circle (4pt) node[lightgray!#4!black,align=center,above=7pt,font=\small]{Regex\\Library};
  \end{tikzpicture}
}
% Jun 12 2018  Add TypeOf  #4671
% Sep 20 2018  Add match type #4964
% ECOOP  Jan 12 → Apr 1  2019, rejected
% OOPSLA Mai 15 → Aug 1  2020, rejected
% PLDI   Nov 20 → Feb 25 2021, rejected
% OOPSLA Apr 16 → Jul 2  2021, rejected
% POPL   Jul 7  → Sep 28 2021, accepted

@@

\begin{slide}{Example 1: Database queries (2016)}
\begin{center}
\includegraphics[width=0.3\textwidth]{figures/spark.pdf}
\end{center}
\end{slide}

\begin{slide}{Example 1: Database queries (2016)}
Spark APIs are practically untyped:
\begin{code}
class DataFrame {
  /** Inner equi-join with another DataFrame on the given column.
   *  The join column will only appear once in the output. */
  def join(right: DataFrame, column: String): DataFrame
}
\end{code}
\pause
Revisited with type-level programming:
\begin{code}
class DF[X] {
  def join[Y](df: DF[Y], col: String): DF[col+(X-col)++(Y-col)]
}
\end{code}
\end{slide}

\begin{slide}{Example 2: Regular expressions (2022)}
Regular expressions in Scala's standard library:
\begin{code}
val rational = new Regex("(\\d+)\\.?(\\d+)?")
rational.unapply("3.1415"): Option[Seq[String]]
\end{code}
\pause
Revisited with type-level programming:
\begin{code}
rational.unapply("3.1415"): Option[(String, Option[String])]
\end{code}
\pause
\begin{code}
class Regex(pattern: String) {
  def unapply(s: String): Option[GroupsOf[@pattern@]]
}
\end{code}
\end{slide}

\begin{slide}{What changed between 2016 and 2022?}
\begin{tikzpicture}
  \draw[black, thick, -Triangle] (0,0) -- (10cm,0);
  \foreach \x in {0,1,...,6}
  \draw[black] (\x*1.4 cm,3pt) -- (\x*1.4 cm,-3pt);

  \foreach \x/\descr in {0/2016,1/2017,2/2018,3/2019,4/2020,5/2021,6/2022}
  \node[black, font=\scriptsize, text height=1.75ex,
  text depth=.5ex] at (\x*1.4,-.3) {$\descr$};

  \fill[implColor] (0.25, 0.5) circle (4pt) node[black,align=center,above=7pt,font=\small]{DataFrame\\Library};

  \fill[mtpeColor] (6.1*1.4, 0.5) circle (4pt) node[black,align=center,above=7pt,font=\small]{Regex\\Library};
\end{tikzpicture}
\end{slide}

\begin{slide}{What changed between 2016 and 2022?}
Example 1 uses a hack, implicits:
\begin{itemize}
  \item convoluted to use
  \item slow to compile
\end{itemize}

Example 2 uses a new language construct, match types, one of the main contributions of this thesis.
\end{slide}

@@

\begin{slide}{Timeline of my PhD}
\timeline{0}{0}{0}{0}{0}
\end{slide}

\begin{slide}{Part I: Generalized Singleton Types}
\timeline{80}{0}{80}{80}{80}
\end{slide}

\begin{slide}{Singleton types}
Scala supports a few singleton types:
\begin{itemize}
  \item |x.type|, the type of the variable |x| (since forever)
  \item |42|, the type of the integer literal |42| (since 2016)
  \item |+|, the type of integer addition (since 2020)
\end{itemize}
\pause
Question: How much of the language can we represent in types?
\end{slide}

\begin{slide}{Generalized singleton types: A proposal}
Our proposal consists of 3 changes:
\begin{enumerate}
  \item new types for if-then-else, pattern matching, constructors, functions calls
  \item "precise mode" of type inference
  \item type evaluation, during subtyping
\end{enumerate}
\end{slide}

\begin{slide}{New types}
What's the type of | if (x == 0) "zero" else "one"|?
\begin{itemize}
  \pause\item |String| ~(Scala 2)
  \pause\item |"zero"|~{\footnotesize\texttt\textbar}~|"one"| ~(Scala 3)
  \pause\item |If[x.type == 0,|| "zero", "one"]| ~(proposed)
  \pause\item |{ if (x == 0) "zero" else "one" }| ~(proposed, syntactic sugar)
\end{itemize}
\pause
Similarly, we add new types for other constructs: |Match[]|, |New[]| and |Call[]|, and the corresponding syntactic sugar.
\end{slide}

\begin{slide}{"Precise mode" of type inference}
We need two modes of type inference, for backwards compatibility.

\vspace{5pt}
\begin{overprint}
\onslide<1>
\begin{code}
@\phantom{dependent}@
def int2str(x: Int) =
  if (x == 0) "zero" else "one"
\end{code}

\onslide<2>
\begin{code}
@\phantom{dependent}@
def int2str(x: Int): String =
  if (x == 0) "zero" else "one"
\end{code}

\onslide<3>
\begin{code}
dependent
def int2str(x: Int)
  : { if (x @\tp{}== 0@) "zero" @\tp{}else@ "one" }
  =   if (x == 0) "zero" else "one"
\end{code}

\onslide<4-5>
\begin{code}
dependent
def int2str(x: Int)
  : If[x.type @\tp{}== 0@, "zero", "one"]
  =   if (x == 0) "zero" else "one"
\end{code}

\onslide<6>
\begin{code}
dependent
def int2str(x: Int)
  : If[x.type @\tp{}== 0@, "zero", String]
  =   if (x == 0) "zero" else readString()
\end{code}

\onslide<7>
\begin{code}
dependent
def int2str(x: Int)
  : { if (x @\tp{}== 0@) "zero" @\tp{}else@ (_: String) }
  =   if (x == 0) "zero" else readString()
\end{code}
\end{overprint}
\medskip
\pause\pause\pause\pause

The "precise mode" is best effort: it lifts whatever possible to the type level and approximates the rest.
\end{slide}

\begin{slide}{List concatenation with generalized singleton types}
\begin{code}
dependent def concat(xs: List, ys: List) <: List =
  xs match
    case x :: xs => x :: concat(xs, ys)
    case Nil => ys
\end{code}
\pause
\begin{overprint}
\onslide<1-2>
\begin{code}
dependent val l1 = "A" :: Nil
dependent val l2 = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { "A" :: "B" :: Nil }
\end{code}

\onslide<3>
\begin{code}
val l1: List     = "A" :: Nil
dependent val l2 = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { concat(l1, l2) }
\end{code}

\onslide<4>
\begin{code}
dependent val l1 = "A" :: Nil
val l2: List     = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { "A" :: l2 }
\end{code}

\end{overprint}
\end{slide}

\begin{slide}{Type evaluation}
During subtyping, we evaluate both sides:
\begin{code}
A <: B  iff  eval(A) <: eval(B)
\end{code}
\pause

Straightforward for if-then-else:
\begin{code}
eval(If[true, A, B]) = A
eval(If[false, A, B]) = B
\end{code}
\pause

More interesting for pattern matching: we "desugar" pattern matching expressions into if-then-else \& type tests.
\pause
\smallskip

We evaluate type tests using subtyping and type disjointness:
\begin{code}
eval(@\tp{}x.asInstanceOf@[Foo]) = true   iff  x.type <: Foo
eval(@\tp{}x.asInstanceOf@[Foo]) = false  iff  disj(x.type, Foo)
\end{code}
\end{slide}

\begin{slide}{Type evaluation: pattern matching example}
\begin{overprint}
\onslide<1>
\begin{code}
dependent val foo(x: Any) =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<2>
\begin{code}
dependent val foo(x: Any): {
  x match
    case s: String => @\tp{}s@
    @\tp{}case i: Int@ => @\tp{}i+1@
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<3->
\begin{code}
dependent val foo(x: Any): {
  if (x.isInstanceOf[String]) x.asInstanceOf[String]
  @\tp{}else if @(@\tp{}x.isInstanceOf@[@\tp{}Int@]) @\tp{}x.asInstanceOf@[Int] @\tp{}+ 1@
  @\tp{}else throw new@ MatchError()
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\end{overprint}
\bigskip
\begin{overprint}
\onslide<4>
|foo(readString())|
\onslide<5>
|foo(readString()): { foo(_: String) }|
\onslide<6>
|foo(readString()): { (_: String).asInstanceOf[String] }|
\onslide<7>
|foo(readString()): { (_: String) }|
\onslide<8>
|foo(readString()): String|
\end{overprint}
\end{slide}

\begin{slide}{Generalized Singletons: Conclusion}
TODO
\end{slide}

% Termination
% Virtual dispatch

\begin{slide}{Part II: Match Types}
\timeline{80}{80}{0}{80}{80}
\end{slide}

\begin{slide}{Overview}
\begin{code}
type Elem[X] = X match
  case String => Char
  case List[t] => Elem[t]
  case Any => X
\end{code}
\end{slide}

\begin{slide}{Disjointness}
|Elem[Seq[Int]]| is stuck
\end{slide}

\begin{slide}{Relating match terms and types: a defining property}
|Elem[Seq[Int]]| is stuck
\end{slide}

\begin{slide}{Formalizing Match Types}
System FM is an extension of System \Fsub with
\begin{itemize}
  \item pattern matching
  \item opaque classes
  \item match types
\end{itemize}
Design goals:
\begin{itemize}
  \item explain the essence of match types
  \item give us confidence in our design
  \item be as simple as possible
\end{itemize}
\end{slide}

\begin{slide}{System FM is parametric}
System FM is parametrized by a set of externally defined classes:

FM(C, $\Psi$, $\Xi$)
\begin{itemize}
  \item $\C \hspace{2.3pt} \coloneqq$ set of classes
  \item $\Psi \coloneqq$ class inheritance relation
  \item $\Xi \hspace{1.1pt} \coloneqq$ class disjointness relation
\end{itemize}

For example:

|class A|                    -> C={A}, $\Psi$={}, $\Xi$={}
|class A; class B extends A| -> C={A, B}, $\Psi$={(A, B)}, $\Xi$={}
|class A; class D|           -> C={A, D}, $\Psi$={}, $\Xi$={(A, D)}
|class A; trait T|           -> C={A, T}, $\Psi$={}, $\Xi$={}
\end{slide}

\begin{slide}{System FM's syntax}
\includegraphics[width=\textwidth]{figures/FMSyntax.pdf}
\end{slide}

\begin{slide}{System FM's disjointness relation}
\includegraphics[width=\textwidth]{figures/FMDisjointness.pdf}
\end{slide}

\begin{slide}{System FM's new subtyping rules}
\includegraphics[width=\textwidth]{figures/FMSubtyping.pdf}
\end{slide}

\begin{slide}{System FM's type safety}
theorems

two versions of the proof, pen and paper and Coq
\end{slide}

\begin{slide}{Structure of the type safety proof}
\includegraphics[width=\textwidth]{figures/FMProofStructure.pdf}
\end{slide}

\begin{slide}{Limitations}
- type binders
- higher-kinded types
- recursion
\end{slide}

\begin{slide}{Empty Types}
\begin{code}
type M[X] = X match
  case Int => String
  case String => Int
class C:
  type X
  def f(bad: M[X & String]): Int = bad
class D extends C:
  type X = Int
\end{code}
\end{slide}

\begin{slide}{Variance}
\begin{code}
F[+T]: T1 <: T2 => F[T1] <: F[T2]
G[-T]: T1 <: T2 => G[T1] >: G[T2]

case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{code}
\end{slide}

\begin{slide}{Match Types, Conclusion}
TODO
\end{slide}

\begin{slide}{Type-Safe Regular Expressions}
\timeline{80}{80}{80}{0}{80}
\end{slide}

\begin{slide}{Type-Safe Regular Expressions (2022)}
\begin{code}
val rational = new Regex("(\\d+)\\.?(\\d+)?")
rational.unapply("3.1415"): Option[(String, Option[String])]
\end{code}
\pause
\begin{code}
type GroupsOf[P <: String] = ...

\end{code}
\end{slide}

% \appendix
\end{document}

### 2022-03-18 symposium.tex
Slides

@@

\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline (an average of +0.3~ms per regex).
This corresponds to the cost of match type reduction.
The implicit-based variant adds another 21~ms (+0.1~ms per regex), which corresponds to the cost of implicit resolution.
Both variants are thus relatively cheap, which validates their utility for practical uses.
When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of type checking \citep[§ 2.11.3]{petrashko2017design}.

>>>>

% TODO: consider inline functions as a 3rd alternative

@@

\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline's total compilation time (an average of +0.3~ms per regex).
This corresponds to the cost of match type reduction.
The implicit-based variant adds another 21~ms (+0.1~ms per regex), which corresponds to the cost of implicit resolution.
Both variants are thus relatively cheap, which validates their utility for practical uses.
When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of type checking \citep[§ 2.11.3]{petrashko2017design}.

### 2022-03-18 thesis.tex
Draft submission
Update symposium.tex
Slides

Scala's implicits allow us to define type-level operations for heterogeneous lists.
We develop the example presented in \Cref{chap:introduction} by looking into the remove operation on |HList|s.
Remove operation takes as argument an element and a list, and returns that list with the first occurrence of the element removed.
This operation should yield an error if the element is not part of the list.

@@

@@

This chapter is inspired by Andrea Veneziano's semester project entitled "Strongly-typed regular expressions in Dotty".
In his project, Andrea implemented a regex library in two different flavors, one based on match types and the other based on generalized singleton types, with the objective to compare the two approaches in terms of expressivity and ease of use.
The present chapter is a reboot of the match-type side of that project.

\input{symposium.tex}

>>>>

Scala's implicits allow us to define type-level operations for heterogeneous lists.
We develop the example presented in \Cref{chap:introduction} by looking into the remove operation on |HList|s.
The remove operation takes as argument an element and a list, and returns that list with the first occurrence of the element removed.
This operation should yield an error if the element is not part of the list.

@@

% TODO: Add a paragraph about the evolution of the "rule for picking a most specific alternative" during implicit resolution, see https://dotty.epfl.ch/docs/reference/changed-features/implicit-resolution.html

@@

This chapter is inspired by Andrea Veneziano's semester project entitled "Strongly-typed regular expressions in Dotty".
In his project, Andrea implemented a regex library in two different flavors, one based on match types and the other based on generalized singleton types, with the objective to compare the two approaches in terms of expressivity and ease of use.
The present chapter is a reboot of the match-type side of that project.
We plan to submit this work to the upcoming Scala Symposium.

\newcommand\thesisOnly[1]{#1}
\newcommand\paperOnly[1]{}
\input{symposium.tex}
